Leadership across cultures and contexts: an exploratory analysis

### 2020-06-22 leadership_across_cultures_contexts.Rmd
Converting to html
Reduced a few words

Our approach was exploratory. Whereas previously we used the ethnographic record to conduct *a priori* tests of leadership theories [@garfield_evolutionary_2019], here we aim to let the ethnographic record of leadership "speak for itself." We did not have *a priori* hypotheses to test.

@@

We use the database of `r nrow(all_data)` ethnographic paragraphs (termed text records) on leadership, described in @garfield_evolutionary_2019 (https://doi.org/10.5281/zenodo.2541999), to code entirely new variables on the functions and qualities of leaders and the costs and benefits of leadership and followership. The text records were extracted from the 60-culture Probability Sample Files (PSF) of the electronic Human Relations Area Files (eHRAF) [@naroll_proposed_1967]. The PSF, which comprises over `r trunc(sum(leader_cult['ehraf_pages'])/1000)`K pages of digitized text, aims to mitigate Galton's problem of cultural non-independence by randomly sampling of one culture with high quality ethnographic coverage from each of 60 regions.

@@

Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgeable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019]. The computational services model emphasizes mating benefits, however, which appear in `r matingbenefit`% of cultures, below our arbitrary 60% threshold for "universality" (Figure \@ref(fig:varsupportcb)). But there was evidence for the *Social, reproductive, and materials success* feature, which included leaders in polygynous unions, having a high quality spouse, and having many children, in more cultures (`r reprosuccess`%) than any other feature (Figure \@ref(fig:featuressupportplot)). Interestingly, the computational services model highlights women's family leadership role, and the estimated marginal mean for the *Social, reproductive, and materials success* feature was higher for females leaders than male leaders (Figure S\@ref(fig:heatmapfeaturesex)), with the caveat that there were few text records on female leaders in our sample.

@@

Coercive authority is a common leadership dimension in the ethnographic record and for which evidence does not vary by any of our four factors, yet did not quite make the 60% cutoff to be included on our list of candidate universals. This dimension features in several evolutionary models of leadership involving dominance hierarchies [@barkow_darwin_1989; @tiger_imperial_1971; @henrich_evolution_2001; @cheng_dominance_2019; @van_vugt_dual_2019] and the role of punishment in promoting cooperation [e.g., @marlowe_more_2008; @henrich_markets_2010; @henrich_costly_2006; @boyd_punishment_1992]. Indeed, we found that coercive authority among leaders was associated with enforcing punishment, aggressiveness, being feared, and also with controlling economics, which evokes "resource holding power", a concept from the behavioral ecology of dominance hierarchies (Figure \@ref(fig:elasticcoercive)). This passage describing the authority of Trobriand chiefs is illustrative [@oj29-007, pp. 64-65]:

@@

These four factors were confounded. Evidence for leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013; @venkataraman_hunter-gatherer_2017]. Evidence for supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories (Figures \@ref(fig:grouptype) and S\@ref(fig:regionsubsistence)). These patterns support perspectives suggesting residential mobility and subsistence intensification are associated with increasing complexity of sociopolitical organization [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014; @kantner_identifying_2010]. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese by @az02-001[p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in *Coding example* section of SI). One interpretation of this trend is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes.

@@

The clearest pattern to emerge from analyses of systematic variation is that, not too surprisingly, many dimensions of leadership are context-specific: aggressiveness and bravery were associated with military leadership, supernatural qualities and ritual functions were associated with religious leadership, organizing cooperation and being experience or accomplished were associated with economic-group leadership, and providing counsel and older age were associated with kin-group leadership (Figure \@ref(fig:heatmapgroups)). (Six leadership dimensions varied by continental region; Figure \@ref(fig:heatmapregion)). We do not have clear theoretical interpretations of this variation, so we treat region simply as a control variable.) These results underscore recent trends in evolutionary approaches to leadership that emphasize the context-dependency of leader emergence and evolutionary selection pressures [@smith_leadership_2020; @smith_leadership_2016; @hagen_leadership_2019].

>>>>

Our approach was exploratory. Whereas previously we used the ethnographic record to conduct *a priori* tests of leadership theories [@garfield_evolutionary_2019], here we aim to let the ethnographic record of leadership "speak for itself."

@@

We use the database of `r nrow(all_data)` ethnographic paragraphs (termed text records) on leadership, described in @garfield_evolutionary_2019 (https://doi.org/10.5281/zenodo.2541999), to code entirely new variables on the functions and qualities of leaders and the costs and benefits of leadership and followership. The text records were extracted from the 60-culture Probability Sample Files (PSF) of the electronic Human Relations Area Files (eHRAF) [@naroll_proposed_1967]. The PSF, which comprises over `r trunc(sum(leader_cult['ehraf_pages'])/1000)`K pages of digitized text, aims to mitigate Galton's problem of cultural non-independence by randomly sampling one culture with high quality ethnographic coverage from each of 60 regions.

@@

Our candidate universal dimensions appear in the computational services model developed by @hagen_leadership_2019, in which knowledgeable and intelligent individuals provide computational (cognitive) services, such as conflict resolution and strategizing, to group members in exchange for a variety of fitness benefits [see also @garfield_evolutionary_2019]. This model emphasizes mating benefits, however, which appear in `r matingbenefit`% of cultures, below our arbitrary 60% threshold for "universality" (Figure \@ref(fig:varsupportcb)). But there was evidence for the *Social, reproductive, and materials success* feature, which included leaders in polygynous unions, having a high quality spouse, and having many children, in more cultures (`r reprosuccess`%) than any other feature (Figure \@ref(fig:featuressupportplot)). Interestingly, the computational services model highlights women's family leadership role, and there was more evidence for the *Social, reproductive, and materials success* feature for females leaders than male leaders (Figure S\@ref(fig:heatmapfeaturesex)), with the caveat that there were few text records on female leaders in our sample.

@@

Coercive authority is a common leadership dimension in the ethnographic record and for which evidence does not vary by any of our four factors, yet did not quite make the 60% cutoff to be included on our list of candidate universals. This dimension features in several evolutionary models of leadership involving dominance hierarchies [@barkow_darwin_1989; @tiger_imperial_1971; @henrich_evolution_2001; @cheng_dominance_2019; @van_vugt_dual_2019] and the role of punishment in promoting cooperation [e.g., @marlowe_more_2008; @henrich_markets_2010; @henrich_costly_2006; @boyd_punishment_1992]. Indeed, we found that coercive authority among leaders was associated with enforcing punishment, aggressiveness, being feared, and also with controlling economics, which evokes "resource holding power", a concept from the behavioral ecology of dominance hierarchies (Figure \@ref(fig:elasticcoercive)). The authority of Trobriand chiefs is illustrative [@oj29-007, pp. 64-65]:

@@

These four factors were confounded. Evidence for leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013; @venkataraman_hunter-gatherer_2017]. Evidence for supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories (Figures \@ref(fig:grouptype) and S\@ref(fig:regionsubsistence)). These patterns support perspectives suggesting residential mobility and subsistence intensification are associated with increasing complexity of sociopolitical organization [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014]. One interpretation of this trend is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes.

@@

The clearest pattern to emerge from analyses of systematic variation is that, not too surprisingly, many dimensions of leadership are context-specific: aggressiveness and bravery were associated with military leadership, supernatural qualities and ritual functions were associated with religious leadership, organizing cooperation and being experience or accomplished were associated with economic-group leadership, and providing counsel and older age were associated with kin-group leadership (Figure \@ref(fig:heatmapgroups)). (Six leadership dimensions varied by continental region; Figure \@ref(fig:heatmapregion). We do not have clear theoretical interpretations of this variation, so we treat region simply as a control variable.) These results underscore recent trends in evolutionary approaches to leadership that emphasize the context-dependency of leader emergence and evolutionary selection pressures [@smith_leadership_2020; @smith_leadership_2016; @hagen_leadership_2019].
